---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# splintr

Natural cubic splines with interpretable intercepts: 'centres' a basis generated using `splines::ns()` on a specified x-value. When used in a model formula, this allows the model intercept to be interpreted with respect to that central x-value, rather than with respect to the x-value of the first `splines::ns()` knot.

## Installation

You can install splintr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("simisc/splintr")
```

## Example

```{r}
library(broom)   # tidy(), augment()
library(knitr)   # kable()
library(splines) # ns()
library(splintr) # splintr()
library(ggplot2) # ggplot()
theme_set(theme_minimal())
```

Using `splines::ns()` in a model formula as below, the model intercept represents the estimated value of `weight` at the first boundary knot, i.e. when `height` takes its minimum value of 58.

```{r example1}
kable(tidy(fm0 <- lm(weight ~ ns(height, df = 3), data = women), quick = TRUE))
attr(ns(women$height, df = 3), "Boundary.knots")
predict(fm0, newdata = data.frame(height = 0))
predict(fm0, newdata = data.frame(height = 58))
ggplot(augment(fm0, data = women), aes(x = height)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = 58, y = coef(fm0)[1]), col = "red")
```

Centring the predictor does not change the model intercept.

```{r example2}
women$height_centred <- women$height - mean(women$height)
kable(tidy(fm1 <- lm(weight ~ ns(height_centred, df = 3), data = women), quick = TRUE))
attr(ns(women$height_centred, df = 3), "Boundary.knots")
predict(fm1, newdata = data.frame(height_centred = 0))
predict(fm1, newdata = data.frame(height_centred = -7))
ggplot(augment(fm1, data = women), aes(x = height_centred)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = -7, y = coef(fm1)[1]), col = "red")
```

Using `splintr()` instead, the intercept representes the estimated value of `weight` when the predictor `height_centred` takes a value of zero.

```{r example3}
kable(tidy(fm2 <- lm(weight ~ splintr(height_centred, df = 3), data = women), quick = TRUE))
attr(splintr(women$height_centred, df = 3), "Boundary.knots")
predict(fm2, newdata = data.frame(height_centred = 0))
predict(fm2, newdata = data.frame(height_centred = -7))
ggplot(augment(fm2, data = women), aes(x = height_centred)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = 0, y = coef(fm2)[1]), col = "red")
```

Alternatively, an arbitrary "centre" can be specified directly in the `splintr()` call.

```{r example4}
x_centre = 68.45
kable(tidy(fm3 <- lm(weight ~ splintr(height, df = 3, centre = x_centre), data = women), quick = TRUE))
attr(splintr(women$height, df = 3, centre = x_centre), "Boundary.knots")
predict(fm3, newdata = data.frame(height = x_centre))
ggplot(augment(fm2, data = women), aes(x = height)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = x_centre, y = coef(fm3)[1]), col = "red")
```

The four models fit identically:

```{r compare}
kable(t(rbind(
  fm0 = glance(fm0),
  fm1 = glance(fm1),
  fm2 = glance(fm2),
  fm3 = glance(fm3)
)))
```
