---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# splintr

Natural cubic splines with interpretable intercepts: 'centres' a basis generated using `splines::ns()` on a specified x-value. When used in a model formula, this allows the model intercept to be interpreted with respect to that central x-value, rather than with respect to the x-value of the first `splines::ns()` knot.

## Installation

You can install splintr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("simisc/splintr")
```

## Example

Using `splines::ns()` in a model formula as below, the model intercept represents the estimated value of `weight` at the first boundary knot, i.e. when `height` takes its minimum value of 58. Centring the predictor does not change the model intercept.

```{r}
library(splines)
library(broom)
tidy(fm1 <- lm(weight ~ ns(height, df = 5), data = women))
attr(ns(women$height, df = 5), "Boundary.knots")
predict(fm1, newdata = data.frame(height = 58))
women$height_centred <- women$height - mean(women$height)
tidy(fm2 <- lm(weight ~ ns(height_centred, df = 5), data = women))
attr(ns(women$height_centred, df = 5), "Boundary.knots")
predict(fm2, newdata = data.frame(height_centred = -7))
```

Using `splintr()` instead, the intercept representes the estimated value of `weight` when the predictor `height_centred` takes a value of zero. Alternatively, a new centre can be specified directly in the `splintr()` call.

```{r}
library(splintr)
tidy(fm3 <- lm(weight ~ splintr(height_centred, df = 5), data = women))
attr(splintr(women$height_centred, df = 5), "Boundary.knots")
predict(fm3, newdata = data.frame(height_centred = 0))
tidy(fm4 <- lm(weight ~ splintr(height, df = 5, centre = 63.4), data = women))
attr(splintr(women$height, df = 5, centre = 63.4), "Boundary.knots")
predict(fm4, newdata = data.frame(height = 63.4))
```
