---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# splintr

[![Build Status](https://travis-ci.com/simisc/splintr.svg?branch=master)](https://travis-ci.com/simisc/splintr)
[![DOI](https://zenodo.org/badge/141533742.svg)](https://zenodo.org/badge/latestdoi/141533742)

Natural cubic splines with interpretable intercepts: 'centres' a basis generated using `splines::ns()` on a specified x-value. When used in a model formula, this allows the model intercept to be interpreted with respect to that central x-value, rather than with respect to the x-value of the first `splines::ns()` knot.

## Installation

You can install splintr from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("simisc/splintr")
```

## Example

```{r message=FALSE, warning=FALSE}
library(broom)   # tidy(), augment()
library(knitr)   # kable()
library(splines) # ns()
library(splintr) # splintr()
library(ggplot2) # ggplot()
library(dplyr)   # %>%
theme_set(theme_minimal())
```

```{r data}
women <- women %>%
  mutate(height_centred = height - mean(height))
women %>%
  head() %>%
  kable()
```

Using `splines::ns()` in a model formula as below, the model intercept represents the estimated value of `weight` at the first boundary knot, i.e. when `height` takes its minimum value of 58.

```{r example1}
fit0 <- lm(weight ~ ns(height, df = 3), data = women)
fit0 %>%
  tidy(quick = TRUE) %>%
  kable()
(knots <- ns(women$height, df = 3) %>%
  attr("Boundary.knots"))
predict(fit0, newdata = data.frame(height = 0))
predict(fit0, newdata = data.frame(height = knots[1]))
augment(fit0, data = women) %>%
  ggplot(aes(x = height)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = knots[1], y = coef(fit0)[1]), col = "red", size = 2)
```

Centring the predictor does not change the model intercept.

```{r example2}
fit1 <- lm(weight ~ ns(height_centred, df = 3), data = women)
fit1 %>%
  tidy(quick = TRUE) %>%
  kable()
(knots <- ns(women$height_centred, df = 3) %>%
  attr("Boundary.knots"))
predict(fit1, newdata = data.frame(height_centred = 0))
predict(fit1, newdata = data.frame(height_centred = knots[1]))
augment(fit1, data = women) %>%
  ggplot(aes(x = height_centred)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = knots[1], y = coef(fit1)[1]), col = "red", size = 2)
```

Using `splintr()` instead, the intercept representes the estimated value of `weight` when the predictor `height_centred` takes a value of zero.

```{r example3}
fit2 <- lm(weight ~ splintr(height_centred, df = 3), data = women)
fit2 %>%
  tidy(quick = TRUE) %>%
  kable()
(knots <- splintr(women$height_centred, df = 3) %>%
  attr("Boundary.knots"))
predict(fit2, newdata = data.frame(height_centred = 0))
augment(fit2, data = women) %>%
  ggplot(aes(x = height_centred)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = 0, y = coef(fit2)[1]), col = "red", size = 2)
```

Alternatively, an arbitrary "centre" can be specified directly in the `splintr()` call.

```{r example4}
x_centre = 68.45
fit3 <- lm(weight ~ splintr(height, df = 3, centre = x_centre), data = women)
fit3 %>%
  tidy(quick = TRUE) %>%
  kable()
(knots <- splintr(women$height, df = 3, centre = x_centre) %>%
  attr("Boundary.knots"))
predict(fit3, newdata = data.frame(height = x_centre))
augment(fit3, data = women) %>%
  ggplot(aes(x = height)) +
  geom_point(aes(y = weight)) +
  geom_line(aes(y = .fitted), col = "blue") +
  geom_point(data = NULL, aes(x = x_centre, y = coef(fit3)[1]), col = "red", size = 2)
```

The four models fit identically:

```{r compare}
rbind(fit0 = glance(fit0),
      fit1 = glance(fit1),
      fit2 = glance(fit2),
      fit3 = glance(fit3)) %>%
  t() %>%
  kable()
```
